---
title: "`rsyncrosim`: introduction to `ST-Sim`"
output: rmarkdown::html_vignette
# output: pdf_document
vignette: >
 %\VignetteIndexEntry{`rsyncrosim`: introduction to `ST-Sim`}
 %\VignetteEngine{knitr::rmarkdown}
 %\usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

This vignette will cover how to develop and run spatially-explicit, stochastic state-and-transition simulation models of landscape change using the `rsyncrosim` package within the <a href="https://syncrosim.com/" target="_blank">SyncroSim</a> software framework. For an overview of <a href="https://syncrosim.com/" target="_blank">SyncroSim</a> and `rsyncrosim`, as well as a basic usage tutorial for `rsyncrosim`, see the <a href="https://syncrosim.github.io/rsyncrosim/articles/a01_rsyncrosim_vignette_basic.html#setup-1" target="_blank">Introduction to `rsyncrosim`</a> vignette.  

## SyncroSim Package: `stsim`

To demonstrate how to create and run spatially-explicit, stochastic state-and-transition simulation models (STSMs) of landscape change, we will be using the <a href="https://github.com/ApexRMS/stsim" target="_blank">stsim</a> SyncroSim package. STSMs are well-suited for integrating uncertainty into model projections, and have been applied to a variety of landscapes and management questions. In this `stsim` example below, we will model changes in forest cover types under two scenarios: no harvest within a landscape, and harvest of 20 acres per year. To do this, we will set a number of model parameters describing how many timesteps and iterations the model will run for, transition types and their probabilities, transition targets (particularly, for the harvest transition type), etc. Next, we will provide `stsim` with a set of initial conditions that describe the starting landscape at time 0, and specify parameters for the model's output. After running both scenarios, we will view the output data in tabular form and as a raster layer. Spatial and non-spatial examples of this exercise are provided.   

For more details on `stsim`, consult the <a href="https://docs.stsim.net/" target="_blank">ST-Sim online documentation</a>.

## Setup

### Install SyncroSim

Before using `rsyncrosim` you will first need to <a href="https://syncrosim.com/download/" target="_blank">download and install</a> the SyncroSim software. Versions of SyncroSim exist for both Windows and Linux.

### Installing and loading R packages

You will need to install the `rsyncrosim` R package, either using <a href="https://cran.r-project.org/" target="_blank">CRAN</a> or from the `rsyncrosim` <a href="https://github.com/syncrosim/rsyncrosim/releases/" target="_blank">GitHub repository</a>. Versions of `rsyncrosim` are available for both Windows and Linux. You may need to install the `raster` and `rgdal` packages from CRAN as well.

In a new R script, load the necessary packages. This includes the `rsyncrosim` and `raster` R packages. 

```{r load packages, message = FALSE}
# Load R packages
library(rsyncrosim)  # package for working with SyncroSim
library(raster)      # package for working with raster data
```

### Connecting R to SyncroSim using `session()`

Finish setting up the R environment for the `rsyncrosim` workflow by creating a SyncroSim Session object. Use the `session()` function to connect R to your installed copy of the SyncroSim software.

```{r load session not run, warning = FALSE, eval = FALSE}
mySession <- session("path/to/install_folder")      # Create a Session based SyncroSim install folder
mySession <- session()                              # Using default install folder (Windows only)
mySession                                           # Displays the Session object
```

```{r load session run, warning = FALSE, echo = FALSE}
# Results of this code shown for above
mySession <- session()                              # Using default install folder (Windows only)
mySession                                           # Displays the Session object
```

Use the `version()` function to ensure you are using the latest version of SyncroSim.

```{r check version, warning = FALSE}
version(mySession)
```

### Installing SyncroSim packages using `installPackage()`

Install `stsim` using the `rsyncrosim` function `installPackage()`. This function takes a package name as input and then queries the SyncroSim package server for the specified package.

```{r remove package from server, warning=FALSE, include=FALSE}
installedPackages <- package()
if (is.element(
  "stsim", installedPackages$name)) uninstallPackage(
    "stsim", force = TRUE)
```

```{r add package from server, warning=FALSE}
# Install stsim
installPackage("stsim")
```

`stsim` should now be included in the package list returned by the `package()` function in `rsyncrosim`:

```{r check packages not run, warning=FALSE, eval=FALSE}
# Get list of installed packages
package()
```

```{r check packages run, warning=FALSE, echo=FALSE}
installedPackages <- package()
stsim_pkg <- installedPackages[installedPackages$name == "stsim", ]
row.names(stsim_pkg) <- NULL
stsim_pkg
```


## Create a modeling workflow


When creating a new modeling workflow from scratch, we need to create objects of the following scopes:

* <a href="https://docs.syncrosim.com/how_to_guides/library_overview.html" target="_blank">Library</a>
* <a href="https://docs.syncrosim.com/how_to_guides/library_overview.html" target="_blank">Projects</a>
* <a href="https://docs.syncrosim.com/how_to_guides/library_overview.html" target="_blank">Scenarios</a>

For more information on these scopes, see the 

<a href="https://syncrosim.github.io/rsyncrosim/articles/a01_rsyncrosim_vignette_basic.html" target="_blank">Introduction to `rsyncrosim`</a> vignette.

### Set up Library, Project, and Scenario

```{r create Library, warning = FALSE}
# Create a new Library
myLibrary <- ssimLibrary(name = "stsimLibrary.ssim",
                         session = mySession,
                         package = "stsim",
                         overwrite = TRUE)

# Open the default Project
myProject <- project(ssimObject = myLibrary, project = "Definitions")

# Create a new Scenario (associated with the default Project)
myScenario <- scenario(ssimObject = myProject, scenario = "My spatial scenario")
```

### View model inputs using `datasheet()`

View the Datasheets associated with your new Project and Scenario using the `datasheet()` function from `rsyncrosim`.

```{r view datasheet list, warning = FALSE}
# View all Datasheets associated with a Library, Project, or Scenario
datasheet_list <- datasheet(myScenario)
head(datasheet_list)
tail(datasheet_list)
```

From this list of Datasheets, we can check which Datasheets specific to the `stsim` package we would like to modify. This will include a number of Initial Conditions Datasheets, Output Datasheets, and a Run Control Datasheet. For more information about all `stsim` Datasheets, see <a href="https://docs.stsim.net/reference/properties.html" target="_blank">the online documentation</a>.

### Configure model inputs using `datasheet()` and `addRow()`

Now, we will add some values to these Datasheets so we can run our models. 

Inputs for `stsim` come in two forms: Scenario inputs (which *can* vary by Scenario) and Project inputs (which *must* be fixed across all Scenarios). In general most inputs are Scenario inputs; Project inputs are typically reserved for values that must be shared by all Scenarios (e.g. constants, shared lookup values). We refer to Project input Datasheets as *Project-scoped Datasheets* and Scenario input Datasheets as *Scenario-scoped Datasheets*. We will start by retrieving some Project-scoped input Datasheets to add and edit their values.  

#### Project-scoped Datasheets

**Terminology**: The Project-scoped Datasheet called `Terminology` specifies terms used across all Scenarios in the same Project.

```{r load datasheet, warning = FALSE}
# Load the Terminology Datasheet to a new R data frame 
terminology <- datasheet(myProject, name = "stsim_Terminology")

# Check the columns of the Terminology data frame
str(terminology)
```

We can change the terminology of the `StateLabelX` and `AmountUnits` columns in this Datasheet, and then save those changes back to the SyncroSim Library file.

```{r save terminology, warning = FALSE}
# Edit the values of the StateLabelX and AmountUnits columns
terminology$AmountUnits <- "Hectares"
terminology$StateLabelX <- "Forest Type"

# Saves edits as a SyncroSim Datasheet
saveDatasheet(myProject, terminology, "stsim_Terminology") 
```

Similarly, we can edit other Project-scoped Datasheets for 'stsim'. 

**Stratum**:  Primary Strata in the model
```{r save stratum, warning = FALSE}
# Load a copy of the Stratum Datasheet.
# To load an empty copy of this Datasheet, specify the argument empty = TRUE in the datasheet() function.
stratum <- datasheet(myProject, "stsim_Stratum", empty = TRUE)

# Use the addRow() to add a value to the stratum  data frame
stratum <- addRow(stratum, "Entire Forest")

# Save edits as a SyncroSim Datasheet
saveDatasheet(myProject, stratum, "stsim_Stratum", force = TRUE)
```

**StateLabelX**: First dimension of labels for State Classes

It is also possible to add values to a Datasheet without loading it into R. Below we create a vector of `forestTypes` and add this to the `stsim_StateLabelX` Datasheet as a data frame using `saveDatasheet()`.
```{r save StateLabelX, warning = FALSE}
# Create a vector containing the State Class labels 
forestTypes <- c("Coniferous", "Deciduous", "Mixed")

# Add values as a data frame to a SyncroSim Datasheet
saveDatasheet(myProject, 
              data.frame(Name = forestTypes), 
              "stsim_StateLabelX", 
              force = TRUE)
```


**StateLabelY**: Second dimension of labels for State Classes
```{r save StateLabelY, warning = FALSE}
# Add values as a data frame directly to an stsim Datasheet
saveDatasheet(myProject, 
              data.frame(Name = c("All")), 
              "stsim_StateLabelY", 
              force = TRUE)
```

**State Classes**: Combine `StateLabelX` and `StateLabelY`, and assign each class a unique name and ID
```{r save State Classes, warning = FALSE}
# Create a new R data frame containing the names of the State Classes and their corresponding data
stateClasses <- data.frame(Name = forestTypes)
stateClasses$StateLabelXID <- stateClasses$Name
stateClasses$StateLabelYID <- "All"
stateClasses$ID <- c(1, 2, 3)

# Save stateClasses R data frame to a SyncroSim Datasheet
saveDatasheet(myProject, stateClasses, "stsim_StateClass", force = TRUE)
```

**Transition Types**: Assign a unique name and ID to each type of transition in our model.
```{r save transitionTypes, warning = FALSE}
# Create an R data frame containing transition type data 
transitionTypes <- data.frame(Name = c("Fire", "Harvest", "Succession"), 
                              ID = c(1, 2, 3))

# Save transitionTypes R data frame to a SyncroSim Datasheet
saveDatasheet(myProject, transitionTypes, "stsim_TransitionType", force = TRUE)
```

**Transition Groups**: Create Transition Groups identical to the Types
```{r save transitionGroups, warning = FALSE}
# Create an R data frame containing a column of transition type names 
transitionGroups <- data.frame(Name = c("Fire", "Harvest", "Succession"))

# Save transitionGroups R data frame to a SyncroSim Datasheet
saveDatasheet(myProject, transitionGroups, "TransitionGroup", force = T)
```

**Transition Types by Groups**: Assign each Type to its Group
```{r save transitionTypesGroups, warning = FALSE}
# Create an R data frame that contains Transition Type Group names
transitionTypesGroups <- data.frame(TransitionTypeID = transitionTypes$Name,
                                    TransitionGroupID = transitionGroups$Name)

# Save transitionTypesGroups R data frame to a SyncroSim Datasheet
saveDatasheet(myProject, 
              transitionTypesGroups, 
              "TransitionTypeGroup", 
              force = T)
```

**Ages**: Define the basic parameters to control the age reporting in the model
```{r save ages, warning = FALSE}
# Define values for age reporting
ageFrequency <- 1
ageMax <- 101
ageGroups <- c(20, 40, 60, 80, 100)

# Add values as R data frames to the appropriate SyncroSim Datasheet
saveDatasheet(myProject, 
              data.frame(Frequency = ageFrequency, MaximumAge = ageMax),
              "AgeType", 
              force = TRUE)
saveDatasheet(myProject, 
              data.frame(MaximumAge = ageGroups), 
              "stsim_AgeGroup", 
              force = TRUE)
```

#### Scenario-scoped Datasheets

Now that we have defined all our Project-scoped input Datasheets, we can move on to specifying Scenario-specific model inputs. We begin by using the `scenario` function to create a new Scenario in our Project.

```{r}
# Create a new SyncroSim Scenario
myScenario <- scenario(myProject, "No Harvest")
```

Once again we can use the `datasheet()` function (with `summary=TRUE`) to display all the Scenario-scoped Datasheets.

```{r subset datasheet by scenario}
# Subset the full Datasheet list to show only Scenario-scoped Datasheets
scenario_datasheet_list <- subset(datasheet(myScenario, summary = TRUE),
                                  scope == "scenario")

head(scenario_datasheet_list)
tail(scenario_datasheet_list)
```

We can now use the `datasheet()` function to retrieve, one at a time, each of our Scenario-scoped Datasheets from our Library.

**Run Control**: Define the length of the run and whether or not it is a spatial run (requires spatial inputs to be set, see below). Here we make the run spatial.
```{r save runControl, warning = FALSE}
# Create an R data frame specifying to run the simulation for 7 realizations and 10 timesteps
runControl <- data.frame(MaximumIteration = 7,
                         MinimumTimestep = 0,
                         MaximumTimestep = 10,
                         isSpatial = TRUE)

# Save transitionTypesGroups R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, runControl, "stsim_RunControl")
```

**Deterministic Transitions**: Define transitions that take place in the absence of probabilistic transitions. Here we also set the age boundaries for each State Class.
```{r save dTransitions, warning = FALSE}
# Load  an empty Deterministic Transitions Datasheet to a new R data frame
dTransitions <- datasheet(myScenario, 
                          "stsim_DeterministicTransition", 
                          optional = T, 
                          empty = T)

# Add all Deterministic Transitions to the R data frame
dTransitions <- addRow(dTransitions, data.frame(
  StateClassIDSource = "Coniferous",
  StateClassIDDest = "Coniferous",
  AgeMin = 21,
  Location = "C1"))
dTransitions <- addRow(dTransitions, data.frame(
  StateClassIDSource = "Deciduous",
  StateClassIDDest = "Deciduous",
  Location = "A1"))
dTransitions <- addRow(dTransitions, data.frame(
  StateClassIDSource = "Mixed",
  StateClassIDDest = "Mixed",
  AgeMin = 11,
  Location = "B1"))

# Save dTransitions R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, dTransitions, "stsim_DeterministicTransition")
```

**Probabilistic Transitions**: Define transitions between State Classes and assigns a probability to each.
```{r save pTransitions, warning = FALSE}
# Load  an empty Probabilistic Transitions Datasheet to a new R data frame
pTransitions <- datasheet(myScenario, "stsim_Transition", optional = T,
                          empty = T)

# Add all Probabilistic Transitions to the R data frame
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Coniferous", 
  StateClassIDDest = "Deciduous", 
  TransitionTypeID = "Fire", 
  Probability = 0.01))
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Coniferous",
  StateClassIDDest = "Deciduous", 
  TransitionTypeID = "Harvest", 
  Probability = 1, 
  AgeMin = 40))
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Deciduous",
  StateClassIDDest = "Deciduous", 
  TransitionTypeID = "Fire", 
  Probability = 0.002))
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Deciduous",
  StateClassIDDest = "Mixed", 
  TransitionTypeID = "Succession", 
  Probability = 0.1, 
  AgeMin = 10))
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Mixed", 
  StateClassIDDest = "Deciduous", 
  TransitionTypeID = "Fire", 
  Probability = 0.005))
pTransitions <- addRow(pTransitions, data.frame(
  StateClassIDSource = "Mixed", 
  StateClassIDDest = "Coniferous",
  TransitionTypeID = "Succession", 
  Probability = 0.1, 
  AgeMin = 20))

# Save pTransitions R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, pTransitions, "stsim_Transition")
```

**Initial Conditions**: Set the starting conditions of the model at time 0. There are two options for setting initial conditions: either spatial or non-spatial. In this example we will use spatial initial conditions; however, we also demonstrate how to set initial conditions non-spatially below.

* ***Option 1 - Spatial***. Let's first take a look at our rasters. We will retrieve the sample raster files (in GeoTIFF format) provided with the `rsyncrosim` package. 

```{r echo = FALSE}
# Load sample .tif files
stratumTif <- "C:/gitprojects/rsyncrosim/vignettes/initial-stratum.tif"
sclassTif <- "C:/gitprojects/rsyncrosim/vignettes/initial-sclass.tif"
ageTif <- "C:/gitprojects/rsyncrosim/vignettes/initial-age.tif"
```

```{r eval = FALSE}
# Load sample .tif files
stratumTif <- "initial-stratum.tif"
sclassTif <- "initial-sclass.tif"
ageTif <- "initial-age.tif"
```


```{r plot initial conditions rasters, warning = FALSE, message = FALSE, fig.align = "center", fig.dim = c(5,5)}
# Create raster layers from the .tif files
rStratum <- raster(stratumTif)
rSclass <- raster(sclassTif)
rAge <- raster(ageTif)

# Plot raster layers
plot(rStratum)
plot(rSclass)
plot(rAge)
```

We can add these rasters as model inputs using the `stsim_InitialConditionsSpatial` Datasheet.

```{r save ICSpatial, warning = FALSE}
# Create an R list of the input raster layers
ICSpatial <- list(StratumFileName = stratumTif, 
                  StateClassFileName = sclassTif, 
                  AgeFileName = ageTif)

# Save initialConditionsSpatial R list to a SyncroSim Datasheet
saveDatasheet(myScenario, ICSpatial, "stsim_InitialConditionsSpatial")
```

* ***Option 2 - Non-spatial***. The second option is to set the proportions of each class, making this a non-spatial parameterization. To do so, we use the `stsim_InitialConditionsNonSpatial` and `stsim_InitialConditionsNonSpatialDistribution` datasheets:
```{r save ICNonSpatial, warning = FALSE}
# Create non-spatial initial conditions data and add it to an R data frame
ICNonSpatial <- data.frame(TotalAmount = 100, 
                           NumCells = 100, 
                           CalcFromDist = F)

# Save the ICNonSpatial R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, ICNonSpatial, "stsim_InitialConditionsNonSpatial")
```


```{r save ICNonSpatialDistribution, warning = FALSE}
# Create non-spatial initial conditions distribution data and add it to an R data frame
ICNonSpatialDistribution <- data.frame(StratumID = "Entire Forest", 
                                       StateClassID = "Coniferous", 
                                       RelativeAmount = 1)

# Save the ICNonSpatial R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, ICNonSpatialDistribution,
              "stsim_InitialConditionsNonSpatialDistribution")
```

**Transition Targets**: Define targets, in units of area, to be reached by the allocation procedure within SyncroSim.
```{r save transitionTarget, warning = FALSE}
# Set the transition target for harvest to 0
saveDatasheet(myScenario, 
              data.frame(TransitionGroupID = "Harvest", 
                         Amount = 0),
              "stsim_TransitionTarget")
```

**Output Options**: Regulate the model outputs and determine the frequency at which syncrosim saves the model outputs.
```{r save outputOptionsSpatial, warning = FALSE}
# Create output options for spatial model and add it to an R data frame
outputOptionsSpatial <- data.frame(
  RasterOutputSC = T, RasterOutputSCTimesteps = 1,
  RasterOutputTR = T, RasterOutputTRTimesteps = 1,
  RasterOutputAge = T, RasterOutputAgeTimesteps = 1
)

# Save the outputOptionsSpatial R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, outputOptionsSpatial, "stsim_OutputOptionsSpatial")

# Create output options for non-spatial model and add it to an R data frame
outputOptionsNonSpatial <- data.frame(
  SummaryOutputSC = T, SummaryOutputSCTimesteps = 1,
  SummaryOutputTR = T, SummaryOutputTRTimesteps = 1
)

# Save the outputOptionsNonSpatial R data frame to a SyncroSim Datasheet
saveDatasheet(myScenario, outputOptionsNonSpatial, "stsim_OutputOptions")
```

We are done parameterizing our simple "No Harvest" Scenario. Let's now define a new Scenario that implements forest harvesting. Below, we create a second "Harvest" Scenario that is a copy of the first Scenario, but with a harvest level of 20 acres/year.

```{r save harvest transitionTarget, warning = FALSE}
# Create a copy of the no harvest scenario (i.e myScenario) and name it myScenarioHarvest
myScenarioHarvest <- scenario(myProject, 
                              scenario = "Harvest", 
                              sourceScenario = myScenario)

# Set the transition target for harvest to 20 acres/year
saveDatasheet(myScenarioHarvest, data.frame(TransitionGroupID = "Harvest", 
                                            Amount = 20), 
              "stsim_TransitionTarget")
```

We can display the harvest levels for both scenarios.

```{r view transitionTarget datasheets, warning = FALSE}
# View the transition targets for the Harvest and No Harvest scenarios
datasheet(myProject, scenario = c("Harvest", "No Harvest"), 
          name = "stsim_TransitionTarget")
```

## Run Scenarios

### Setting run parameters with `run()`

We will now run both Scenarios using the `run()` function in `rsyncrosim`. Running a Scenario generates a corresponding new child Scenario, called a *Results Scenario*, which contains the results of the run along with a snapshot of all the model inputs. If we have a large modeling workflow and we want to parallelize the run using multiprocessing, we can set the `jobs` argument to be a value greater than one.

```{r run scenarios, warning = FALSE}
# Run both Scenarios
myResultScenario <- run(myProject, scenario = c("Harvest", "No Harvest"), 
                     jobs = 7, summary = TRUE)

```

## View results

The next step is to view the output Datasheets added to the Result Scenario when it was run. To look at the results we first need to retrieve the unique `scenarioId` for each child *Result Scenario*.

```{r}
# Retrieve Scenario IDs
resultIDNoHarvest <- subset(myResultScenario, 
                            ParentId == scenarioId(myScenario))$ScenarioId
resultIDHarvest <- subset(myResultScenario, 
                          ParentId == scenarioId(myScenarioHarvest))$ScenarioId
```

We can now retrieve tabular output regarding the projected State Class over time (for both scenarios combined) from the `stsim_OutputStratumState` Datasheet.

```{r}
# Retrieve output projected State Class for both Scenarios in tabular form
outputStratumState <- datasheet(
  myProject, 
  scenario = c(resultIDNoHarvest, resultIDHarvest), 
  name = "stsim_OutputStratumState")
```

Finally, we can get the State Class raster output using the `datasheetRaster()` function (here for the Harvest scenario only). 

```{r view myRastersTimestep5, warning = FALSE, fig.align = TRUE, fig.dim = c(5,5)}
# Retrieve the output State Class raster for the Harvest scenario at timestep 5
myRastersTimestep5 <- datasheetRaster(myProject, 
                                      scenario = resultIDHarvest, 
                                      "stsim_OutputSpatialState", 
                                      timestep = 5)
myRastersTimestep5

# Plot raster for the first realization of timestep 5
plot(myRastersTimestep5[[1]])
```
