---
title: "Installing beta release"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing beta release}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

Follow these instructions to install the `rsyncrosim` beta release from GitHub.

1. Open RStudio (or any R IDE)

2. Make sure that you do not have another version of the `rsyncrosim` library loaded in your R session. You can detach a loaded library using the following code:

```{r step 2, eval=FALSE}
detach("package:rsyncrosim", unload=TRUE)
```

3. Use the `install.packages()` function to install `rsyncrosim` from a website URL. Use the following code to install version 1.3.0 from the `rsyncrosim` GitHub repository:

```{r step 3, eval=FALSE}
install.packages("https://github.com/syncrosim/rsyncrosim/releases/download/1.3.0/rsyncrosim_1.3.0.tar.gz", repo=NULL)
```

4. You many need to restart your R environment to get rid of any environmental conflicts. To do this easily, use the following code:

```{r step 4, eval=FALSE}
.rs.restartR()
```

5. Load the `rsyncrosim` library.

```{r step 5, eval=FALSE}
library(rsyncrosim)
```



