---
title: "Updating rsyncrosim scripts to version 2"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating rsyncrosim scripts to version 2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

The introduction of <a href="https://syncrosim.com/studio/" target="_blank">SyncroSim Studio</a> brought some structural changes and new functionalities, which get reflected in the `rsyncrosim` package. Use the guide below to update your `rsyncrosim v1` scripts to `rsyncrosim v2`.

*Note*: <a href="https://syncrosim.com/studio-download/" target="_blank">SyncroSim v3.0.9</a> or higher is required to use `rsyncrosim v2.0.1` or higher. 

## Updated functions

To install and uninstall packages from your SyncroSim installation, use:

```{r install packages, eval=FALSE}
installPackage() # Installs the selected package(s) to SyncroSim
uninstallPackage() # Uninstalls the selected package(s)
```

Now you can use multiple SyncroSim packages in a single library. To add and remove packages from your library, use:

```{r add package, eval=FALSE}
addPackage() # Adds package(s) to the library
removePackage() # Removes package(s) from the library. Note that any associated datasheets will be removed as well.
```

And also for that reason, the package addons have been converted to standalone packages, or incorporated into the base packages. Therefore, the following functions no longer exist:

```{r deprecated, eval=FALSE}
addon()
disableAddon()
enableAddon()
```

## Library

`info(ssimLibrary)` no longer shows the *Package name*, *Current package version* and *Minimum package version*, and the Input and Output folder paths have been combined into the *Data folder*.

the `package` argument in the `ssimLibrary(package)` function has been renamed to `packages`

```{r library, eval=FALSE}
myLibrary <- ssimLibrary(packages = c())
```

The `addon` and `template` arguments have also been removed.

## Dependencies

To view the existing dependencies for a scenario, use:

```{r view dependencies, eval=FALSE}
dependency(myScenario)
```

To set dependedncies for a scenario, use:

```{r set dependencies, eval=FALSE}
dependency(myScenario) <- c("Scenario 1", "Scenario 2")
```

Note that this will remove any previously set dependencies.
Also, the order in which the dependencies are listed can be important. In this example, `Scenario 1` takes precedence over `Scenario 2`.


## Multiprocessing

The `run()` function no longer has the `jobs` argument. Use the `core_Multiprocessing` datasheet to set the number of cores to use. Note that this will affect every scenario run.


## Variable naming

The **ID** in variable names have been changed to **Id**. It is generally safe to substitute `ID` for `Id` throughout your script, unless that affects any variables not related to SyncroSim.

## stsimsf

The ST-Sim Stock & Flow package, `stsimsf` have been incorporated into ST-Sim. Substitute all `stsimsf_` datasheets throghout your script for `stsim_`.
The only exception is the `stsimsf_OutputOptions` that is now called `stsim_OutputOptionsStockFlow`.

## Other removed functions

```{r other removed, eval=FALSE}
datasheetRaster() # previously deprecated function. Use datasheetSpatRaster() instead.

addBreakpoint()
breakpoint()
deleteBreakpoint()

ssimUpdate()
updatePackage()
```

## Other NEW functions
```{r new functions, eval=FALSE}
createCondaEnv(pkgs = c()) # creates the conda environment for the selecte package(s)

deleteLibrary() # Note this is irreversible

chart() # Create or load a SyncroSim chart. Explore other charting functions with ?chart
```

