---
title: "Updating rsyncrosim scripts to version 2"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating rsyncrosim scripts to version 2}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

The introduction of <a href="https://syncrosim.com/studio/" target="_blank">SyncroSim Studio</a> brought some structural changes and new features, which are reflected in the `rsyncrosim` package. Use the guide below to update your `rsyncrosim` version 1 scripts to be compatible with version 2.

*Note*: <a href="https://syncrosim.com/studio-download/" target="_blank">SyncroSim v3.0.9</a> or higher is required to use `rsyncrosim v2.0.0` or higher. 

## Updated functions

### Installing/uninstalling packages

Previously, the `addPackage` and `removePackage` functions were used to install and uninstall packages from SyncroSim. Now to install and uninstall packages from SyncroSim, use:

```{r install packages, eval=FALSE}
installPackage()   # Installs the selected package(s) to SyncroSim
uninstallPackage() # Uninstalls the selected package(s) from SyncroSim
```

### Adding/removing packages

Now you can use multiple SyncroSim packages in a single library. To add and remove packages from your library, use:

```{r add package, eval=FALSE}
addPackage()    # Adds package(s) to the library
removePackage() # Removes package(s) from the library.
```
Note that when you remove a package from your library any associated datasheets will be removed as well.

You can also have multiple versions of a package installed, and choose which package version to use within a library. Use the `versions` argument in the `installPackage()` and `addPackage()` functions.

## Deprecated functions

In SyncroSim 3, the concept of addon packages no longer exists. All addon packages have either been converted to standalone packages (e.g., `burnP3PlusPrometheus`), and can be added to a library without having to load a base package first, or incorporated directly into the base packages (e.g., `stsimsf` is now part of `stsim`). Therefore, the following functions have been removed:

```{r deprecated, eval=FALSE}
addon()
disableAddon()
enableAddon()
```

Since you can install multiple versions of a package, the `updatePackage()` function has been removed. To update (or downgrade) the package version on a library, use the `versions` argument in the `addPackages()` function:
```{r changeVersion, eval=FALSE}
# Change package version
addPackage(myLibrary, packages = "stsim", versions = "4.0.0")
addPackage(myLibrary, packages = "stsim", versions = "4.0.1")
```

Any functions depending on the deprecated `raster` package have been removed, including the `datasheetRaster()` function. Use `datasheetSpatRaster()` instead.

Other deprecated functions include:

```{r otherDeprecated, eval=FALSE}
addBreakpoint()
breakpoint()
deleteBreakpoint()

ssimUpdate()
```

## Library

Now that addon packages no longer exist and SyncroSim libraries can support multiple packages, the following changes have been made to the `ssimLibrary()` function:

* The `addon` argument has been removed.
* The `template` argument has been removed. This is a temporary change, but in the meantime you can download template libraries from <a href="https://cloud.syncrosim.com/" target="_blank">SyncroSim Cloud</a>.
* The `package` argument has been renamed to `packages` (plural).

```{r library, eval=FALSE}
# rsyncrosim version 1:
myLibrary <- ssimLibrary(package = "PackageName", addon = "AddonName")

# rsyncrosim version 2:
myLibrary <- ssimLibrary(packages = c("stsim", "helloworldSpatial"))
```
```{r load session run, warning = FALSE, echo = FALSE}
# to show the packages added, plus create scenarios to show below
library(rsyncrosim)
mySession <- session()
myLibrary <- ssimLibrary("Changes to rsyncrosim v2.ssim", packages=c("stsim", "helloworldSpatial"))
scenario1 <- scenario(myLibrary, scenario = "Scenario 1")
scenario2 <- scenario(myLibrary, scenario = "Scenario 2")
scenario3 <- scenario(myLibrary, scenario = "Scenario 3")
myScenario <- scenario(myLibrary, "my Scenario")
dependency(myScenario) <- "Scenario 1"
```

`info(ssimLibrary)` no longer shows the *Package name*, *Current package version* and *Minimum package version*, and the Input and Output folder paths have been combined into the *Data folder*. See the <a href="https://syncrosim.github.io/rsyncrosim/articles/a01_rsyncrosim_vignette_basic.html#access-model-metadata" taret="_blank">Introduction to rsyncrosim</a> for details.


## Dependencies

Scenario dependencies are now set using the `<-` operator in `rsyncrosim`. See below for examples on how the usage of the `dependency()` function has changed.

To view the existing dependencies for a scenario, use:

```{r view dependencies}
dependency(myScenario)
```

To set dependencies for a scenario:

```{r set dependencies, eval=FALSE}
# rsyncrosim version 1:
dependency(myScenario, dependency = c("Scenario 2", "Scenario 3"))

```
```{r dependencies, warning=FALSE}
# rsyncrosim version 2:
dependency(myScenario) <- c("Scenario 2", "Scenario 3")

dependency(myScenario)

```

Note that this will remove any previously set dependencies unless the existing dependencies are also included in the vector.
Also, the order in which the dependencies are listed can be important. In this example, `Scenario 2` takes precedence over `Scenario 3`.


## Multiprocessing

The `run()` function no longer has the `jobs` argument. Use the `core_Multiprocessing` library datasheet to set the number of cores to use. Note that this will affect every scenario run.

```{r multiprocessing, eval=FALSE}
multiprocessing <-
    data.frame(EnableMultiprocessing = TRUE,
               MaximumJobs = 6)
  
saveDatasheet(myLibrary, multiprocessing, "core_Multiprocessing")
```


## Variable naming

The **ID** in the primary key changed to **Id** for all datasheets. That also means that the column names for variables depending on higher-level datasheets also changed their suffixes to **Id**. It is generally safe to substitute `ID` for `Id` throughout your script, but it's recommended to check the datasheet's variable names.

```{r IDs, warning=FALSE}
datasheet(myScenario, "stsim_FlowPathway", includeKey = TRUE, optional = TRUE, empty = TRUE)
```

## Core datasheets

The <a href="https://docs.syncrosim.com/reference/ds_overview.html" target="_blank">system datasheets</a>, previously identified by `corestime_`, are now simply called `core_`.


## Other NEW functions
```{r new functions, eval=FALSE}
createCondaEnv(pkgs = c()) # creates the conda environment for the selected package(s)

deleteLibrary() # Note this is irreversible

chart() # Create or load a SyncroSim chart. Explore other charting functions with ?chart
```

